Compare scANVI and SingleR annotations

### 2025-08-18 analyses/cell-type-neuroblastoma-04/exploratory-notebooks/scanvi-singler-comparison.Rmd
Fix some spelling while I'm here for #1276
substantially simplify code using updated family TSV

# Source additional utilities functions:
# - select_nbatlas_markers()
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

@@

# label mapping
nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)

@@

# sample metadata
sample_metadata <- readr::read_tsv(sample_metadata_file)
```

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "Unknown"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "Unknown", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) |>
  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label)) |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)

# can remove now for space
rm(merged_sce)
```

@@

```{r, fig.width = 12, fig.height = 8}
# define families using the provided map
nbatlas_t_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "T cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_nk_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "NK cell") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

# Note that cDC and Monocyte are not direct labels in the disseminated NBAtlas object,
# but these groupings are used in the paper's marker gene analysis.
nbatlas_cdc_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "cDC") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

nbatlas_monocyte_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Monocyte") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

# we'll use these later but not for this plot
nbatlas_myeloid_celltypes <- nbatlas_label_map |>
  dplyr::filter(NBAtlas_label == "Myeloid") |>
  dplyr::pull(NBAtlas_label_immune_zoom)

umap_direct_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  # shuffle points to ensure order they are plotted is random
  dplyr::sample_frac(1) |> 
  # order celltypes so that families are grouped together for a cleaner legend
  # note this will give warnings if any of those cells aren't in the annotation, but that's fine
  dplyr::mutate(
    label = factor(label), 
    label = forcats::fct_relevel(
      label, 
      "T cell", nbatlas_t_celltypes, 
      "NK cell", nbatlas_nk_celltypes,
      "cDC", nbatlas_cdc_celltypes, 
      "Monocyte", nbatlas_monocyte_celltypes
    )
  )

@@

* Fraction of identical labels
* Fraction of labels that agree at least on the level of their cell type "family," where we define these families based on groups of NBAtlas annotations:
  * T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
  * Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
  * Conventional dendritic cells (cDC): `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
  * Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
  * Myeloid: `Neutrophil` and `Macrophage`
* Fraction of labels that entirely disagree

@@

# create a wide data frame with "family" labels where possible
celltype_family_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  # create label_family grouping
  dplyr::mutate(
    label_family = dplyr::case_when(
      label %in% nbatlas_t_celltypes ~ "T cell",
      label %in% nbatlas_nk_celltypes ~ "NK cell",
      label %in% nbatlas_cdc_celltypes ~ "cDC",
      label %in% nbatlas_monocyte_celltypes ~ "Monocyte",
      label %in% nbatlas_myeloid_celltypes ~ "Myeloid", 
      .default = label
    )
  ) |>
  dplyr::select(-label_recoded, -label) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label_family
  )

>>>>

# Source additional utilities functions:
# - subset_nbatlas_markers()
# - harmonize_celltypes()
# - faceted_umap()
# - generate_dotplot()
source(file.path(module_dir, "scripts", "utils", "celltype-utils.R"))
```

@@

@@

# sample metadata
sample_metadata <- readr::read_tsv(sample_metadata_file)

# label mapping
nbatlas_label_map <- readr::read_tsv(nbatlas_label_map_file)
```

@@

```{r}
celltype_df <- scuttle::makePerCellDF(
  merged_sce,
  use.coldata = c("sample_id", "is_xenograft", "consensus_celltype_annotation"),
  use.dimred = c("UMAP")
) |>
  # there ARE repeated barcodes so we need to keep cell_id around
  tibble::rownames_to_column("cell_id") |>
  dplyr::rename(
    UMAP1 = UMAP.1,
    UMAP2 = UMAP.2,
    consensus_annotation = consensus_celltype_annotation
  ) |>
  dplyr::left_join(validation_df, by = "consensus_annotation") |>
  # recode consensus NAs to "Unknown"
  dplyr::mutate(
    validation_group_annotation = ifelse(
      is.na(validation_group_annotation), "Unknown", validation_group_annotation
    )
  ) |>
  dplyr::select(-consensus_annotation) |>
  # merge in the annotations
  dplyr::left_join(singler_df, by = "cell_id") |>
  dplyr::left_join(scanvi_df, by = "cell_id") |>
  # make it tidy
  tidyr::pivot_longer(
    c(validation_group_annotation, singler, scanvi), 
    names_to = "celltype_method", 
    values_to = "label"
  ) |>
  harmonize_celltypes(label, label_recoded) |>
  dplyr::mutate(label = ifelse(label == "NE", "Neuroendocrine", label)) |>
  tidyr::separate(cell_id, into = c("library_id", "barcode"), remove = FALSE)
```

@@

```{r, fig.width = 12, fig.height = 8}
label_order <- nbatlas_label_map |>
  dplyr::add_count(NBAtlas_family) |>
  dplyr::arrange(desc(n), NBAtlas_family) |>
  dplyr::pull(NBAtlas_label)

umap_direct_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  # shuffle points to ensure order they are plotted is random
  dplyr::sample_frac(1) |> 
  # order celltypes so that families are grouped together for a cleaner legend
  # note this will give warnings if any of those cells aren't in the annotation, but that's fine
  dplyr::mutate(label = factor(label, levels = label_order))

@@

* Fraction of identical labels
* Fraction of labels that agree at least on the level of their cell type "family," where we define these families based on groups of NBAtlas annotations:
  * T cells: `T cell`, `Treg`, `CD4+ T cell`, `CD8+ T cell`
  * Natural killer cells: `NK cell`, `Circulating NK cell`, `Resident NK cell`, `TOX2+/KIT+ NK cell`
  * Conventional dendritic cells (cDC): `cDC`, `cDC1`, `cDC2/DC3`, `Migratory cDC`
  * Monocytes: `Monocyte`, `Classical monocyte`, `Patrolling monocyte`
* Fraction of labels that entirely disagree

@@

# create a wide data frame with "family" labels where possible
celltype_family_df <- celltype_df |>
  dplyr::filter(celltype_method != "validation_group_annotation") |>
  dplyr::left_join(
    nbatlas_label_map, by = c("label" = "NBAtlas_label")
  ) |>
  dplyr::select(cell_id, library_id, celltype_method, label) |>
  tidyr::pivot_wider(
    names_from = celltype_method,
    values_from = label
  )
